Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-04-12 mpaper/paper.tex
Short conclusions and pseudocode!

\documentclass{mpaper}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mathtools}
\usepackage{thmtools, thm-restate}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{tikz}
\usepackage{layouts}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{csvsimple}
\usepackage{pgfplots}
\usepackage{blkarray}

@@

As we want to restrict some parameters (namely, $\bm\lambda$ and the diagonal of
$\mathbf{B}$) to positive values, we express them as exponentials and later
adjust their derivatives accordingly. Specifically, we can set $\lambda_i =
e^{\lambda'_i}$ and optimise $\lambda'_i$ using the chain rule:
\[
  \frac{\partial \mathcal{L}}{\partial \lambda'_i} = e^{\lambda'_i}
  \frac{\partial \mathcal{L}}{\partial \lambda_i}.
\]
This way, we restrict $\lambda_i$ to positive values while allowing $\lambda'_i$
to range over $\mathbb{R}$.

Finally, the parameters can be initialised as follows\footnote{In practice, it
  is often easier to start with a fixed $\mathbf{B} = \mathbf{I}_m$, as random
  values of $\mathbf{B}$ can often result in a near-singular $\bm\Sigma$.}:
\begin{alignat*}{3}
  \mu_i &\sim \mathcal{U}(0, 1) \quad &&\text{for } i = 1, \dots, m, \\
  \lambda_0 &\sim \chi^2_5, && \\
  \lambda_i &\sim \chi^2_1 \quad &&\text{for } i = 1, \dots, d, \\
  \diag(\mathbf{B}) &\sim \chi^2_4, && \\
  \text{the rest of } \mathbf{B} &\sim \mathcal{N}(0, 1). &&
\end{alignat*}
The initialisation of $\bm\mu$ mirrors the initialisation of $\mathbf{r}$ in
previous work by Levine et al. \cite{DBLP:conf/nips/LevinePK11}. In contrast,
while they have constant initial values for $\bm\lambda$, we sample from
$\chi^2$ distributions centred around those values ($5$ for $\lambda_0$ and $1$
for any other $\lambda_i$). The distributions for initial values of $\mathbf{B}$
are simply set to provide a reasonable spread of positive values for the
diagonal, and both positive and negative values for all other entries in the
matrix.

@@

Having proven that differentiation can be moved inside the integral allows us to
use $\nabla\mathcal{L}$ for optimisation, knowing that our approach is sound.
This allows us to implement a simple optimisation algorithm with a constant step
size. The algorithm was implemented in MATLAB and is compatible with the
toolkit\footnote{\url{https://graphics.stanford.edu/projects/gpirl/}} for
evaluating IRL algorithms developed by Levine et al.
\cite{DBLP:conf/nips/LevinePK11}. Note that due to the stochastic nature of the
gradient, dynamic step size algorithms such as AdaGrad
\cite{DBLP:journals/jmlr/DuchiHS11} and AdaDelta
\cite{DBLP:journals/corr/abs-1212-5701} proved ineffective, as spikes in the
gradient estimates would drive the step size down to zero. Similarly, although
we check the $\ell_1$-norm of the difference between old and new parameter
values in order to detect convergence, the algorithm usually only stops when the
maximum number of iterations is exhausted.
% TODO (must-y): An algorithm box / pseudo code outlining the algorithms (with
% links to the many equations) so the reader can easily get an impression of the
% (training) algorithm and the complexity.

@@

The main goal of this section is use experimental evidence in order to show that
the implementation agrees with theory and to investigate its properties such as
convergence and how the learned model changes with varying amounts and types of
data. We begin with a three-state MDP where the agent can deterministically move
from any state to any other state. More formally, we set $\mathcal{S} = \{ s_1,
s_2, s_3 \}$, $\mathcal{A} = \{ a_1, a_2 \}$,
\begin{align*}
  &\mathcal{T}(s_1, a_1, s_2) = 1, \quad \mathcal{T}(s_1, a_2, s_3) = 1, \\
  &\mathcal{T}(s_2, a_1, s_1) = 1, \quad \mathcal{T}(s_1, a_2, s_3) = 1, \\
  &\mathcal{T}(s_3, a_1, s_1) = 1, \quad \mathcal{T}(s_1, a_2, s_2) = 1,
\end{align*}
all other values of $\mathcal{T}$ to zero, and $\gamma = 0.9$. We
also set the inducing points to be equal to the three states in $\mathcal{S}$,
add a single feature $f : \mathcal{S} \to \mathbb{R}$ such that
\[
  f(s_1) = 1, \quad f(s_2) = 2, \quad f(s_3) = 3,
\]
and create two demonstrations $\zeta_1 = \{ (s_1, a_1) \}$ and $\zeta_2 = \{
(s_3, a_2) \}$ that correspond to moving from $s_1$ and from $s_3$ to $s_2$.
Therefore, we would expect the reward of $s_2$ to be higher than the other two
rewards in order to reflect this.

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\columnwidth]{figures/convergence_new}
  \caption{Convergence of $\mathcal{L}$ (at the top) as well as several example
    policies (at the bottom)}
  \label{fig:convergence1}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \begin{tikzpicture}[
    /pgfplots/colormap/violet,
    ellipse2/.style={
      draw,
      circle,
      /utils/exec={
        \pgfplotscolormapdefinemappedcolor{#1}},
      fill=mapped color},
    edge2/.style={
      ultra thick,
      /utils/exec={
        \pgfplotscolormapdefinemappedcolor{#1}},
      draw=mapped color}]
    \csvreader[
    no head,
    column count=11
    ]{data/median_covariance_transformed.csv}{1=\x, 2=\y, 3=\Soneone, 4=\Stwotwo, 5=\Sthreethree, 6=\Stwoone, 7=\Sthreeone, 8=\Sthreetwo}{%
      \node[ellipse2=\Soneone] (s1) at (2*\x, 2*\y) {$s_1$};
      \node[ellipse2=\Stwotwo] (s2) at (2*\x+0.5, 2*\y+1) {$s_2$};
      \node[ellipse2=\Sthreethree] (s3) at (2*\x+1, 2*\y) {$s_3$};
      \draw[edge2=\Stwoone] (s1) -- (s2);
      \draw[edge2=\Sthreeone] (s1) to (s3);
      \draw[edge2=\Sthreetwo] (s2) to (s3);
    }
    \draw [<->] (1.5, 7.5) node (yaxis) [above] {$\zeta_2$}
    |- (8, 1.5) node (xaxis) [right] {$\zeta_1$};
    \foreach \x in {1, 2, 3} {
      \draw[shift={(0.5+2*\x, 1.5)}] (0pt,2pt) -- (0pt,-2pt) node[below] {$\x$};
    }
    \foreach \y in {1, 2, 3} {
      \draw[shift={(1.5, 0.5+2*\y)}] (2pt,0pt) -- (-2pt,0pt) node[left] {$\y$};
    }
    \begin{axis}[
      colormap/violet,
      scale only axis,
      width=0pt,
      height=0pt,
      hide axis,
      colorbar horizontal,
      point meta min=-1,
      point meta max=1,
      xshift=40,
      yshift=30,
      colorbar style={
        width=6.5cm,
        xtick={-1,0,1},
        xlabel=normalised covariance,
      }]
      \addplot [draw=none] coordinates {(0,0)};
    \end{axis}
  \end{tikzpicture}
  \caption{Changes in reward covariances after adding more demonstrations. The
    meaning of external axes is the same as in Figure~\ref{fig:policies}.
    Colours denote reward covariance values, e.g., the colour of node $s_1$
    denotes the variance of $r(s_1)$, and the colour of the edge between nodes
    $s_1$ and $s_2$ denotes the covariance between $r(s_1)$ and $r(s_2)$. The
    colours represent median covariances across ten runs, normalised to the
    interval $[-1, 1]$ while preserving their positivity/negativity.}
  \label{fig:covariance}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \begin{tikzpicture}
    \foreach \i in {0,...,9}
    \node[draw,circle] (v\i) at (360/10 * \i:3cm) {$s_{\i}$};
    \foreach \i in {0,...,9}
    \foreach \j in {0,...,\i}
    \draw[gray] (v\i) -- (v\j);

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figures/boxplots}
  \caption{Box plots of the difference between (absolute values of) reward
    covariances in random and semi-structured scenarios from
    Figure~\ref{fig:cliques}, averaged out over 100 runs and grouped into four
    key categories. `First state' refers to the variance of $r(s_0)$, the reward
    of the state targeted by all demonstrations in the semi-structured case.
    `Other states' are variances of other states of the MDP, i.e., diagonal
    entries of the covariance matrix. `Incident edges' refers to covariances
    between $r(s_0)$ and all other states, i.e., the first row or column of the
    covariance matrix. Finally, `other edges' is the category for the remaining
    covariances between pairs of rewards.}
  \label{fig:covariance_boxplots}
\end{figure}

\section{Conclusions and Further Work} \label{sec:conclusions}
% TODO: write
% Show how we achieved our claimed contributions

\begin{itemize}
\item Why my theoretical contributions are important...
\item Reasonable results with policy convergence.
\item Shown how to eliminate the deterministic training conditional
  assumption---a common weakness of previous approaches.
\item Being able to model rewards using full probability distributions without
  limiting assumptions opens up many new research directions.
\item Determining how reward covariance depends on data would be the next step.
\end{itemize}

% FURTHER WORK

An interesting extension to our work would be to consider IRL in the context
of a reinforcement learning (RL) agent. Suppose we have an agent whose purpose
is to learn optimal behaviour from observing other agents using IRL. It could
then take reward variance estimates into account when choosing what states to
visit next. It would have to handle the balance between exploration and
exploitation similarly to many RL agents, but the information about rewards
would come from observing (presumably near-optimal) behaviour exhibited by other
agents rather than directly from the environment.

@@

Finally, building on the idea that variance estimates can be used to judge
whether the model has learned optimal policy, an interesting question for
MDP (or, perhaps, dynamical systems) research would be: how much does a reward
have to change in order to affect the deterministic policy? A simple answer to
this question would allow us to use variance estimates in order to quantify the
model's confidence regarding optimal behaviour.

@@

\section{Derivatives of the ELBO}

>>>>

\documentclass{mpaper}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mathtools}
\usepackage{thmtools, thm-restate}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{tikz}
\usepackage{layouts}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{csvsimple}
\usepackage{pgfplots}
\usepackage{blkarray}
\usepackage[boxed]{algorithm2e}

@@

As we want to restrict some parameters (namely, $\bm\lambda$ and the diagonal of
$\mathbf{B}$) to positive values, we express them as exponentials and later
adjust their derivatives accordingly. Specifically, we can set $\lambda_i =
e^{\lambda'_i}$ and optimise $\lambda'_i$ using the chain rule:
\begin{equation} \label{eq:transformation}
  \frac{\partial \mathcal{L}}{\partial \lambda'_i} = e^{\lambda'_i}
  \frac{\partial \mathcal{L}}{\partial \lambda_i}.
\end{equation}
This way, we restrict $\lambda_i$ to positive values while allowing $\lambda'_i$
to range over $\mathbb{R}$.

Finally, the parameters can be initialised as follows\footnote{In practice, it
  is often easier to start with a fixed $\mathbf{B} = \mathbf{I}_m$, as random
  values of $\mathbf{B}$ can often result in a near-singular $\bm\Sigma$.}:
%\begin{equation} \label{eq:initialisation}
\begin{alignat}{3}
  \mu_i &\sim \mathcal{U}(0, 1) \quad &&\text{for } i = 1, \dots, m, \label{eq:first_initialisation} \\
  \lambda_0 &\sim \chi^2_5, && \\
  \lambda_i &\sim \chi^2_1 \quad &&\text{for } i = 1, \dots, d, \\
  \diag(\mathbf{B}) &\sim \chi^2_4, && \\
  \text{the rest of } \mathbf{B} &\sim \mathcal{N}(0, 1). \label{eq:last_initialisation} &&
\end{alignat}
%\end{equation}
The initialisation of $\bm\mu$ mirrors the initialisation of $\mathbf{r}$ in
previous work by Levine et al. \cite{DBLP:conf/nips/LevinePK11}. In contrast,
while they have constant initial values for $\bm\lambda$, we sample from
$\chi^2$ distributions centred around those values ($5$ for $\lambda_0$ and $1$
for any other $\lambda_i$). The distributions for initial values of $\mathbf{B}$
are simply set to provide a reasonable spread of positive values for the
diagonal, and both positive and negative values for all other entries in the
matrix.

@@

Having proven that differentiation can be moved inside the integral allows us to
use $\nabla\mathcal{L}$ for optimisation, knowing that our approach is sound.
This allows us to implement a simple optimisation algorithm, the description of
which is in the next section.

@@

The main goal of this section is use experimental evidence to show that the
implementation agrees with theory and to investigate its properties such as
convergence and how the learned model changes with varying amounts and types of
data. We begin with an outline of the implementation in
Algorithm~\ref{alg:optimisation}. The algorithm was implemented in MATLAB and is
compatible with the
toolkit\footnote{\url{https://graphics.stanford.edu/projects/gpirl/}} for
evaluating IRL algorithms developed by Levine et al.
\cite{DBLP:conf/nips/LevinePK11}. We decided to use a constant step size because
the stochastic nature of the gradient makes dynamic step size algorithms such as
AdaGrad \cite{DBLP:journals/jmlr/DuchiHS11} and AdaDelta
\cite{DBLP:journals/corr/abs-1212-5701} ineffective, as spikes in the gradient
estimates drive the step size down to zero. Similarly, although we check for
convergence using the $\ell_\infty$ norm, the algorithm usually only stops when
the maximum number of iterations is exhausted.

\begin{algorithm}
  \SetKwData{parameters}{parameters}
  \SetKwData{stepSize}{stepSize}
  \SetKwData{ttolerance}{tolerance}
  \SetKwData{numIterations}{numIterations}
  \SetKwFunction{estimatedPart}{estimatedPart}
  \SetKwFunction{exactPart}{exactPart}
  \SetKwFunction{pack}{pack}
  \SetKwFunction{unpack}{unpack}
  \SetKwFunction{valueIteration}{valueIteration}
  \SetKwFunction{transformGradient}{transformGradient}
  \SetKwProg{Function}{Function}{}{end}
  \KwData{number of samples $S$, constant step size \stepSize, convergence
    criteria \ttolerance and \numIterations}
  \KwResult{optimised \parameters}
  initialise \parameters as in Equations~\eqref{eq:first_initialisation}--\eqref{eq:last_initialisation}\;
  \For{$i \gets 1$ \KwTo \numIterations}{
    $\{ \mathbf{u}_s \}_{s=1}^S \sim \mathcal{N}(\parameters.\bm\mu, \parameters.\bm\Sigma)$\;
    $\nabla\mathcal{L} \gets \exactPart{\parameters} + \frac{1}{S} \sum_{s=1}^S\estimatedPart{\parameters, $\mathbf{u}_s$}$\;
    $\mathbf{v} \gets$ \pack{\parameters}\;
    $\mathbf{v}' \gets \mathbf{v} + \stepSize \times \transformGradient{$\nabla\mathcal{L}$}$\;
    $\parameters \gets \unpack{$\mathbf{v}'$}$\;
    \lIf{$\lVert \mathbf{v}' - \mathbf{v} \rVert_\infty < \ttolerance$}{break}
  }
  \Function{\estimatedPart{\parameters, $\mathbf{u}$}}{
    $\mathbf{r} \sim \mathcal{N}(\parameters.\mathbf{Su}, \parameters.\bm\Gamma)$\;
    $\V \gets \valueIteration{$\mathbf{r}$}$\;
    \Return{the estimated part of $\nabla\mathcal{L}$ according to $\mathbf{r}$,
      $\V$, and \parameters\;}
  }
  \caption{The optimisation procedure. All parameters are kept in a single variable
    \textsf{parameters}. Converting it to/from vector form and using $\log$,
    $\exp$, and Equation~\eqref{eq:transformation} to handle the optimisation of
    parameters restricted to positive values is handled by functions
    \texttt{pack}, \texttt{unpack}, and \texttt{transformGradient}. Each
    derivative from Appendix~\ref{appendix:derivatives} is split into parts
    outside and inside $\mathbb{E}[\cdot]$ which are handled differently by
    functions \texttt{exactPart} and \texttt{estimatedPart}. The function
    \texttt{valueIteration} simply applies Equation~\eqref{eq:update_rule} until
    convergence.}
  \label{alg:optimisation}
\end{algorithm}

We will first focus on a simple three-state MDP where the agent can
deterministically move from any state to any other state. More formally, we set
$\mathcal{S} = \{s_1, s_2, s_3 \}$, $\mathcal{A} = \{ a_1, a_2 \}$,
\begin{align*}
  &\mathcal{T}(s_1, a_1, s_2) = 1, \quad \mathcal{T}(s_1, a_2, s_3) = 1, \\
  &\mathcal{T}(s_2, a_1, s_1) = 1, \quad \mathcal{T}(s_1, a_2, s_3) = 1, \\
  &\mathcal{T}(s_3, a_1, s_1) = 1, \quad \mathcal{T}(s_1, a_2, s_2) = 1,
\end{align*}
all other values of $\mathcal{T}$ to zero, and $\gamma = 0.9$. We
also set the inducing points to be equal to the three states in $\mathcal{S}$,
add a single feature $f : \mathcal{S} \to \mathbb{R}$ such that
\[
  f(s_1) = 1, \quad f(s_2) = 2, \quad f(s_3) = 3,
\]
and create two demonstrations $\zeta_1 = \{ (s_1, a_1) \}$ and $\zeta_2 = \{
(s_3, a_2) \}$ that correspond to moving from $s_1$ and $s_3$ to $s_2$. With
this setup, we would expect the reward of $s_2$ to be higher than the other two
rewards.

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figures/convergence_new}
  \caption{Convergence of $\mathcal{L}$ (at the top) as well as several example
    policies (at the bottom)}
  \label{fig:convergence1}
\end{figure}

@@

\begin{figure}[!t]
  \centering
  \begin{tikzpicture}[
    /pgfplots/colormap/violet,
    ellipse2/.style={
      draw,
      circle,
      /utils/exec={
        \pgfplotscolormapdefinemappedcolor{#1}},
      fill=mapped color},
    edge2/.style={
      ultra thick,
      /utils/exec={
        \pgfplotscolormapdefinemappedcolor{#1}},
      draw=mapped color}]
    \csvreader[
    no head,
    column count=11
    ]{data/median_covariance_transformed.csv}{1=\x, 2=\y, 3=\Soneone, 4=\Stwotwo, 5=\Sthreethree, 6=\Stwoone, 7=\Sthreeone, 8=\Sthreetwo}{%
      \node[ellipse2=\Soneone] (s1) at (2*\x, 2*\y) {$s_1$};
      \node[ellipse2=\Stwotwo] (s2) at (2*\x+0.5, 2*\y+1) {$s_2$};
      \node[ellipse2=\Sthreethree] (s3) at (2*\x+1, 2*\y) {$s_3$};
      \draw[edge2=\Stwoone] (s1) -- (s2);
      \draw[edge2=\Sthreeone] (s1) to (s3);
      \draw[edge2=\Sthreetwo] (s2) to (s3);
    }
    \draw [<->] (1.5, 7.5) node (yaxis) [above] {$\zeta_2$}
    |- (8, 1.5) node (xaxis) [right] {$\zeta_1$};
    \foreach \x in {1, 2, 3} {
      \draw[shift={(0.5+2*\x, 1.5)}] (0pt,2pt) -- (0pt,-2pt) node[below] {$\x$};
    }
    \foreach \y in {1, 2, 3} {
      \draw[shift={(1.5, 0.5+2*\y)}] (2pt,0pt) -- (-2pt,0pt) node[left] {$\y$};
    }
    \begin{axis}[
      colormap/violet,
      scale only axis,
      width=0pt,
      height=0pt,
      hide axis,
      colorbar horizontal,
      point meta min=-1,
      point meta max=1,
      xshift=40,
      yshift=30,
      colorbar style={
        width=6.5cm,
        xtick={-1,0,1},
        xlabel=normalised covariance,
      }]
      \addplot [draw=none] coordinates {(0,0)};
    \end{axis}
  \end{tikzpicture}
  \caption{Changes in reward covariances after adding more demonstrations. The
    meaning of external axes is the same as in Figure~\ref{fig:policies}.
    Colours denote reward covariance values, e.g., the colour of node $s_1$
    denotes the variance of $r(s_1)$, and the colour of the edge between nodes
    $s_1$ and $s_2$ denotes the covariance between $r(s_1)$ and $r(s_2)$. The
    colours represent median covariances across ten runs, normalised to the
    interval $[-1, 1]$ while preserving their positivity/negativity.}
  \label{fig:covariance}
\end{figure}

@@

% TODO: must shorten something by ~1 line
\begin{figure}[!t]
  \centering
  \begin{tikzpicture}
    \foreach \i in {0,...,9}
    \node[draw,circle] (v\i) at (360/10 * \i:3cm) {$s_{\i}$};
    \foreach \i in {0,...,9}
    \foreach \j in {0,...,\i}
    \draw[gray] (v\i) -- (v\j);

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figures/boxplots}
  \caption{Box plots of the difference between (absolute values of) reward
    covariances in random and semi-structured scenarios from
    Figure~\ref{fig:cliques}, averaged out over 100 runs and grouped into four
    key categories. `First state' refers to the variance of $r(s_0)$, the reward
    of the state targeted by all demonstrations in the semi-structured case.
    `Other states' are variances of other states of the MDP, i.e., diagonal
    entries of the covariance matrix. `Incident edges' refers to covariances
    between $r(s_0)$ and all other states, i.e., the first row or column of the
    covariance matrix. Finally, `other edges' is the category for the remaining
    covariances between pairs of rewards.}
  \label{fig:covariance_boxplots}
\end{figure}
% TODO: probably need to conclude the experiments

\section{Conclusions and Further Work} \label{sec:conclusions}
% Show how we achieved our claimed contributions

In this paper, we showed how VI can be used to solve the IRL problem. We
proved the validity of our approach, establishing subsidiary results such as new
bounds for the linearly solvable MDP model along the way. The implementation of
the model was well-behaved and showed that covariance can indeed be used to
recognise a well-learned model from a model which is less sure about what
policies are optimal.

% FURTHER WORK
Building on the that, an interesting question for MDP (or, perhaps, dynamical
systems) research would be: how much does a reward have to change in order to
affect the deterministic policy? A simple answer to this question would allow us
to use variance estimates in order to quantify the model's confidence regarding
optimal behaviour.

@@

Finally, an interesting extension to our work would be to consider IRL in the
context of a reinforcement learning (RL) agent. Suppose we have an agent whose
purpose is to learn optimal behaviour from observing other agents using IRL. It
could then take reward variance estimates into account when choosing what states
to visit next. It would have to handle the balance between exploration and
exploitation similarly to many RL agents, but the information about rewards
would come from observing (presumably near-optimal) behaviour exhibited by other
agents rather than directly from the environment.

@@

\section{Derivatives of the ELBO} \label{appendix:derivatives}
